---
title: "Data preparation"
output: html_document
date: June 2023
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(rgbif)
library(scrubr)
library(maps)
```

```{r}
source("R/Data_preparation_functions.r")
```

## Polygons of Norwegian lakes
The lake polygons need to be downloaded from here:  https://bird.unit.no/resources/9b27e8f0-55dd-442c-be73-26781dad94c8/content, click on “Innhold”-tab at the bottom of the page and select "Norwegian_lakes.rds".

```{r}
lakes <- readRDS("data/Norwegian_lakes.rds")
```


## Citizen science observations from the Norwegian Species Observation Service

The citizen science observations can be downloaded using the `rgbif` package. We specify the species, country and dataset key to get the relevant subset of observations.

```{r}
myspecies <- c("Salvelinus alpinus", "Esox lucius", "Salmo trutta", "Perca fluviatilis")

artsobs_raw <- rgbif::occ_data(scientificName = myspecies, 
                      hasCoordinate = TRUE, 
                      limit = 20000, 
                      country = "NO", # Only observations in Norway
                      year = "1990,2023",
                      # Norwegian Species Observation Service: 
                      datasetKey = "b124e1e0-4755-430f-9eab-894f25a9b59c")

artsobs_data_list <- vector("list", length(myspecies))
names(artsobs_data_list) <- myspecies
for (s in myspecies) {
  sub_data <- artsobs_raw[[s]]$data
  artsobs_data_list[[s]] <- data.frame(species = s, sub_data)
}

artsobs_raw_joined <- bind_rows(artsobs_data_list)

artsobs_matched <- match_to_lake(artsobs_raw_joined, lakes, max_dist_from_lake = 50)[[1]] 

artsobs <- artsobs_matched %>% 
  mutate(occurenceStatus = as.logical(mapvalues(occurrenceStatus, c("ABSENT", "PRESENT"), c(FALSE,TRUE))))

```

## Survey data

The survey data is downloaded from here: https://gbif.vm.ntnu.no/ipt/resource?r=fish_status_survey_of_nordic_lakes. The downloaded file contains two data sets, one with occurrences and one with events, so we need to merge these. 

```{r}
# Merging occurence and events for the survey data:
occ <- read.table("data/survey_raw/occurrence.txt", 
                  header = TRUE, sep = "\t", quote = "\"'", fill = FALSE) %>% 
  mutate(species = gsub("verbatim scientific name: ", "", taxonRemarks)) %>% 
  filter(species %in% myspecies)

event <- read.table("data/survey_raw/event.txt", 
                    header = TRUE, sep = "\t", quote = "\"'", fill = FALSE)

survey_raw <- merge(occ, event, by = "eventID")

# Should the survey data be matched to lakes?
survey_list <- match_to_lake(survey_raw, lakes, max_dist_from_lake = 50)
survey_matched <- survey_list[[1]]
survey <- survey_matched %>% 
  mutate(occurrenceStatus = as.logical(mapvalues(occurrenceStatus, 
                                c("absent", "present"), c(FALSE,TRUE))))
# TODO: slightly odd that 80% of the observations are removed... Is it the non-norwegian ones?
```

## Saving the data

```{r}
saveRDS(survey, "data/survey_clean_new.rds")
saveRDS(artsobs, "data/artsobs_clean_new.rds")
```


