---
title: "Supporting information for 'The Point Process Framework for Integrated Modelling of Biodiversity Data'"
author: "Kwaku Peprah Adjei, Philip Mostert, Jorge Sicacha Parada, Emma Skarstein, Robert B. O'Hara"
output: 
  pdf_document
header-includes:
  - \geometry{top=1in}
  - \usepackage{titling}
  - \pretitle{\begin{flushleft}\Huge\bfseries}
  - \posttitle{\end{flushleft}}  
  - \preauthor{\begin{flushleft}\Large}
  - \postauthor{\end{flushleft}}  
  - \predate{\begin{flushleft}\large}
  - \postdate{\end{flushleft}}
---

Last updated: `r format(Sys.Date(), "%B %d %Y")` at `r format(Sys.time(), "%H:%M:%S")`.

```{r, echo = FALSE}
# Run model, or load from saved version? (and same for predictions)
run_model <- TRUE
run_prediction <- TRUE
```

```{r, echo = FALSE}
start_time <- Sys.time()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.fullwidth = TRUE)
```

# Setup

These are the packages we will be using:

```{r, loading packages, message = FALSE}
#devtools::install_github("PhilipMostert/PointedSDMs")
library(PointedSDMs) # model fitting
library(ggplot2)     # plotting
library(raster)      # plotting
library(mapproj)     # map options for plotting
library(INLA)        # functions for specifying mesh
library(dplyr)       # data handling
library(sf)          # spatial stuff
library(showtext)    # font for plot
library(patchwork)   # combining figures
```

# Downloading data

Before you run this file, make sure you have the following files in the given locations:

-   `data/environmental_covariates.rds`
    -   should be in the data folder already.
-   `data/Norwegian_lakes.rds`
    -   can be downloaded from <https://bird.unit.no/resources/9b27e8f0-55dd-442c-be73-26781dad94c8/content>, click on “Innhold”-tab at the bottom of the page and select "Norwegian_lakes.rds".
-   `data/artsobs_clean.rds`
    -   created from running the script `R/data_preparation.R`, see further instructions there.
-   `data/survey_clean.rds`
    -   also created from running the script `R/data_preparation.R`, see further instructions there.

# Loading Norway map and making mesh

We begin by making a spatial mesh out of a map of Norway.

```{r, norway map stuff, warning = FALSE, message = FALSE, eval = FALSE, echo = FALSE}
#norway <- ggplot2::map_data("world", region = "Norway(?!:Svalbard)")
#norway <- setdiff(norway, dplyr::filter(norway, subregion == "Jan Mayen"))
#norway_sf <- st_as_sf(norway, coords = c("long", "lat"), crs = proj)
```

```{r, norway map, warning = FALSE, message = FALSE}
proj <- '+proj=utm +zone=32 +datum=WGS84 +units=km +no_defs'#m
norwayfill <- maps::map("world", "norway", fill=TRUE, plot=FALSE, 
                  ylim=c(58,72), xlim=c(4,32))
IDs <- sapply(strsplit(norwayfill$names, ":"), function(x) x[1])
norway.poly <- spTransform(maptools::map2SpatialPolygons(norwayfill, IDs = IDs, 
          proj4string = CRS('+proj=lonlat +zone=32 +datum=WGS84 +units=m +no_defs')), proj)
```

Adjusting the mesh to be coarser is the easiest way to decrease the run-time for the model. With the following mesh, the model estimation takes a long time to complete, but feel free to change the `max.edge` or `cutoff` to get a coarser mesh.

```{r, make small mesh, eval = TRUE, fig.width=8, fig.height=5}
mesh <- inla.mesh.2d(boundary = inla.sp2segment(norway.poly), 
                     cutoff = 0.3 * 30, # smallest allowed distance between points
                     max.edge = c(6, 3) * 30, # decrease this for more int. points
                     #max.edge = c(3, 1),
                     offset = c(1, 1) * 30,
                     crs = st_crs(proj))

ipoints = st_as_sf(fm_int(mesh, norway.poly,  
                 int.args = list(method = 'direct')))

plot(mesh)
```

# Setting up covariate data

Next we load the environmental data, which will be used as covariates in our model.

```{r, loading covariates}
covariates_raw <- readRDS("data/environmental_covariates.RDS")

covariates <- covariates_raw %>% 
  # Log-transform area of lake
  dplyr::mutate(log_area = log(area_km2)) %>% 
  # Log-transform catchment area of lake
  dplyr::mutate(log_catchment = log(catchment_area_km2)) %>% 
  # Remove some uninformative variables
  dplyr::select(-c(ebint, no_vatn_lnr, eb_waterregionID))

# Choose from 
# "decimalLatitude", "decimalLongitude",
# "log_area", "perimeter_m", "distance_to_road", 
# "eurolst_bio10", "catchment_area_km2", "SCI", "HFP"

Use <- c("SCI", 'HFP')

cov_pixel <- SpatialPixelsDataFrame(
  points = covariates[,c("decimalLongitude","decimalLatitude")],
  data = data.frame(covariates[,Use]), 
  proj4string = CRS('+proj=lonlat +zone=32 +datum=WGS84 +units=m +no_defs'), 
  tol = 0.340571)
if (length(Use) == 1) names(cov_pixel@data) <- Use 
# Scale covariates and convert to terra::rast
cov_raster <- terra::project(scale(terra::rast(cov_pixel)), proj)
```

# Observation data

For this model, we have two observation sets, one which is downloaded from GBIF and one that is a survey dataset (see separate document for download instructions).

```{r, loading observation data}
fishes <- c("Esox_lucius", "Perca_fluviatilis", "Salmo_trutta", "Salvelinus_alpinus")

survey <- readRDS("data/survey_clean.rds") %>% 
  filter(species %in% fishes) %>%
  st_as_sf(coords =c("decimalLongitude", "decimalLatitude"), 
           crs = '+proj=lonlat +zone=32 +datum=WGS84 +units=m +no_defs') %>%
  mutate(decimalLongitude = st_coordinates(.)[,1], decimalLatitude = st_coordinates(.)[,2]) %>%
  st_transform(proj)
artsobs <- readRDS("data/artsobs_clean.rds") %>% 
  filter(species %in% fishes) %>%
  st_as_sf(coords =c("decimalLongitude", "decimalLatitude"), 
           crs = '+proj=lonlat +zone=32 +datum=WGS84 +units=m +no_defs') %>%
  mutate(decimalLongitude = st_coordinates(.)[,1], decimalLatitude = st_coordinates(.)[,2]) %>%
  st_transform(proj)
```

We can plot the observed data points:

```{r, plot observations}
showtext_auto()
showtext_opts(dpi = 300)
f1 <- "Open sans"
font_add_google(f1, f1)

norway <- ggplot2::map_data("world", region = "Norway(?!:Svalbard)")
norway <- setdiff(norway, dplyr::filter(norway, subregion == "Jan Mayen"))

p_artsobs <- ggplot(artsobs, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_polygon(data = norway, aes(long, lat, group = group), 
               color="grey80", fill = "grey95") + 
  geom_point(color = "darkorange2", size = 0.5, alpha = 0.3) +
  facet_wrap(~species, nrow = 1, 
             labeller = labeller(species = function(string) sub("_", " ", string))) +
  coord_map() +
  labs(tag = "Citizen science data") +
  theme_minimal() +
  theme(text = element_text(family = f1),
        strip.text = element_text(family = f1, size = 12, face = "italic"),
        plot.tag = element_text(angle = 90, hjust = 0.5),
        plot.tag.position = c(-0.03, 0.45),
        legend.position = "none",
        axis.title = element_blank(),
        axis.text.x = element_blank())

p_survey <- ggplot(survey %>% filter(occurrenceStatus == 1), 
       aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_polygon(data = norway, aes(long, lat, group = group), 
               color="grey80", fill = "grey95") + 
  geom_jitter(color = "darkcyan", size = 0.5, alpha = 0.3) +
  facet_wrap(~species, nrow = 1) +
  coord_map() +
  xlab("Longitude") +
  ylab("Latitude") +
  labs(tag = "Survey data") +
  theme_minimal() +
  theme(text = element_text(family = f1),
        plot.tag = element_text(angle = 90, hjust = 0.5),
        plot.tag.position = c(-0.03, 0.45),
        legend.position = "none",
        axis.title = element_blank(),
        strip.text = element_blank(),
        plot.margin = margin(l = 30))

p_artsobs / p_survey

ggsave("figures/presence_points.pdf", height = 5, width = 8)
ggsave("figures/presence_points.png", height = 5, width = 8)
```

# Separate models for the two datasets

For the presence/absence survey data, we use a Bernoulli distribution, where the presence probability for species $j\in \{\textit{Salmo trutta, Perca fluviatilis, Esox lucius, Salvelinus alpinus}\}$ depends on some covariates $x(s)$, along with a spatial field $\xi_{j}(s)$:

$$
\begin{aligned}
    Y_{PA, j}(s_i) &\sim \text{Bernoulli}(p_{PA, j}(s_i)) \\
    \text{cloglog}(p_{PA, j}(s_i)) &= \alpha_{PA, j} + x(s_i)^T\beta_j + \xi_{j}(s_i).
\end{aligned}
$$

We first prepare the model using the `intModel` function.

```{r set up survey model}
surveyModel <- intModel(
  survey,                              # Survey data
  Boundary = st_as_sf(norway.poly),    # Boundary
  IPS = ipoints,                       # integration points
  spatialCovariates = cov_raster,      # Covariates
  speciesName = "species",             # The column containing species name
  Coordinates = c("decimalLongitude", "decimalLatitude"), # Name of coords
  responsePA = "occurrenceStatus",     # Name of response column
  pointsSpatial = NULL,                # NULL since we use speciesSpatial
  Mesh = mesh,                         # inla mesh object
  Projection = proj                    # CRS for points and covariates
  )                                     

```

We then specify priors for the spatial effects using `.$specifySpatial` and for the intercepts using `.$priorsFixed`.

```{r spatialfields survey}
for (fish in fishes) {
  
  surveyModel$specifySpatial(Species = fish, 
                             prior.range = c(50, 0.01),
                             prior.sigma = c(0.5, 0.01))
  
  surveyModel$priorsFixed(Effect = 'intercept', 
                        Species = fish, 
                        mean.linear = 0, 
                        prec.linear = 1)
}



```

And fit the model using `fitISDM`.

```{r fitISDM survey, warning = FALSE}
surveyFit <- fitISDM(surveyModel, 
                      options = list(num.threads = 2, 
                        control.inla = list(int.strategy = 'eb', 
                                            cmin = 0, 
                                            strategy = 'adaptive'),
                        safe = TRUE,
                        inla.mode = 'experimental'))

summary(surveyFit)
```

The presence-only data is fitted with a Poisson point process model, where the intensity depends on the same covariates $x(s)$ and the same spatial field $\xi_{j}(s)$. In the ISDM, we will add an additional spatial field $\xi_{\text{bias}}(s)$ that is unique to the citizen science data, but shared across all fish species:

$$
\begin{aligned}
    Y_{PO, j}(s_i) &\sim \text{Poisson}(e^{\eta_{PO, j}(s_i)}) \\
    \eta_{PO, j}(s_i) &= \alpha_{PO, j} + x(s)^T\beta_j + \xi_{j}(s_i) + \xi_{\text{bias}}(s_i).
\end{aligned}
$$

```{r setup arts model}
artsobsModel <- intModel(
  artsobs,                             # Citizen science data
  Boundary = st_as_sf(norway.poly),    # Boundary
  IPS = ipoints,                       # integration points
  spatialCovariates = cov_raster,      # Covariates
  speciesName = "species",             # The column containing species name
  Coordinates = c("decimalLongitude", "decimalLatitude"), # Name of coords
  responsePA = "occurrenceStatus",     # Name of response column
  pointsSpatial = NULL,                # NULL since we use speciesSpatial
  Mesh = mesh,                         # inla mesh object
  Projection = proj                    # CRS for points and covariates
  )   
```

```{r spatialfields artsobs}
for (fish in fishes) {
  
  artsobsModel$specifySpatial(Species = fish,
                             prior.range = c(50, 0.01),
                             prior.sigma = c(0.5, 0.01))
  
  artsobsModel$priorsFixed(Effect = 'intercept', 
                        Species = fish, 
                        mean.linear = 0, 
                        prec.linear = 1)
}

```

```{r fitISDM artsobs, warning = FALSE}
artsobsFit <- fitISDM(artsobsModel, 
                      options = list(num.threads = 2, 
                        control.inla = list(int.strategy = 'eb', 
                                            cmin = 0, 
                                            strategy = 'gaussian'),
                        safe = TRUE,
                        inla.mode = 'experimental'))

summary(artsobsFit)
```

# Joint model for four fish species

Now we will fit two IDMS, one with the two datasets combined, and another with an additional spatial field for the citizen science data. Both these models will have four shared fields, shared across the data sets (survey/citizen science), but separate for each fish species. Since we have two data sets and four species, that means that we in total have eight sub-models.

```{r, setting up fish model, warning=FALSE}
ISDMsetup <- intModel(
  survey,                              # Survey data
  artsobs,                             # Citizen science data
  Boundary = st_as_sf(norway.poly),    # Boundary
  IPS = ipoints,                       # integration points
  spatialCovariates = cov_raster,      # Covariates
  speciesName = "species",             # The column containing species name
  speciesSpatial = "copy",             # Copy species fields across data
  Coordinates = c("decimalLongitude", "decimalLatitude"), # Name of coords
  responsePA = "occurrenceStatus",     # Name of response column
  pointsSpatial = NULL,                # NULL since we use speciesSpatial
  Mesh = mesh,                         # inla mesh object
  Projection = proj                    # CRS for points and covariates
  )                   
```

```{r jointModel setup spatial}
for (fish in fishes) {
  
  ISDMsetup$specifySpatial(Species = fish,
                           prior.range = c(50, 0.01),
                           prior.sigma = c(0.5, 0.01))
  
  ISDMsetup$priorsFixed(Effect = 'intercept', 
                        Species = fish, 
                        mean.linear = 0, 
                        prec.linear = 1)
}

```

For the species specific fields, the default in PointedSDMs is that these are allowed to be different up to a scaling factor (named beta in INLA) across the data sets. That means that for instance the trout-specific spatial field for the citizen science data set is equal to beta times the trout-specific spatial field for the survey data. In practice, this has to do with the copy-option in INLA. But in our model we want them to be the same, not to vary by a factor. So we manually change this using the `$changeComponents` function, by setting `hyper = list(beta = list(fixed = TRUE))` for each of the four citizen science fields.

```{r, customizing fish model, warning=FALSE, include = TRUE, eval = TRUE}
ISDMsetup$changeComponents(
  'Perca_fluviatilis_artsobs_spatial(
      main = geometry, 
      copy = "Perca_fluviatilis_survey_spatial", 
      hyper = list(beta = list(fixed = TRUE)))',
  print = FALSE)

ISDMsetup$changeComponents(
  'Salmo_trutta_artsobs_spatial(
      main = geometry, 
      copy = "Salmo_trutta_survey_spatial", 
      hyper = list(beta = list(fixed = TRUE)))',
  print = FALSE)

ISDMsetup$changeComponents(
  'Salvelinus_alpinus_artsobs_spatial(
      main = geometry, 
      copy = "Salvelinus_alpinus_survey_spatial", 
      hyper = list(beta = list(fixed = TRUE)))',
  print = FALSE)

ISDMsetup$changeComponents(
  'Esox_lucius_artsobs_spatial(
      main = geometry, 
      copy = "Esox_lucius_survey_spatial", 
      hyper = list(beta = list(fixed = TRUE)))')
```

We may look at which terms are included in each of the eight sub-models by calling `$updateFormula` with the data sets as the arguments.

```{r, view formulas, warning=FALSE}
ISDMsetup$updateFormula(datasetName = "survey")
ISDMsetup$updateFormula(datasetName = "artsobs")
```

```{r, fitting fish model, warning=FALSE, eval = run_model}
ISDM <- fitISDM(ISDMsetup, 
                      options = list(num.threads = 2, 
                        control.inla = list(int.strategy = 'eb', 
                                            cmin = 0, 
                                            strategy = 'gaussian'),
                        safe = TRUE,
                        inla.mode = 'experimental'))
```

We may then examine the model summary and save the model for future use.

```{r, view and save model, warning=FALSE, eval = run_model}
summary(ISDM)

#saveRDS(ISDM, "results/ISDM.rds")
```

This model has the species specific spatial fields, but we also want a bias field that is shared across the species. We add this using `$addBias`.

```{r, add bias field, warning=FALSE}
ISDMsetup$addBias("artsobs")
ISDMsetup$specifySpatial(Bias = 'artsobs',
                         prior.range = c(50, 0.01),
                         prior.sigma = c(0.5, 0.01))
```

```{r, fitting fish model2, warning=FALSE, eval = run_model}
ISDMbias <- fitISDM(ISDMsetup, 
                      options = list(num.threads = 2, 
                        control.inla = list(int.strategy = 'eb', 
                                            cmin = 0, 
                                            strategy = 'gaussian'),
                        safe = TRUE,
                        inla.mode = 'experimental'))

summary(ISDMbias)
#saveRDS(ISDMbias, "results/ISDMbias.rds")
```

```{r, load model, warning=FALSE, echo = FALSE}
#ISDM <- readRDS("results/ISDMbias.rds")
```

# Predictions and plots

Once the model has been fit, we can look at the predictions from the species-specific shared fields and the bias field.

First we create a plot to show the coefficients and the associated credibility intervals for the four different models considered. The estimates and standard errors for the four models are relatively similar.

```{r, coef plots,fig.width=8, fig.height=5}

makeData <- function(data, name) {
  
 data$summary.fixed$variable <- rownames(data$summary.fixed)
 data$summary.fixed$min <- data$summary.fixed$`0.025quant`
 data$summary.fixed$max <- data$summary.fixed$`0.975quant`
 data$summary.fixed$model <- name
 data$summary.fixed

}

plotData <- rbind(makeData(surveyFit, 'survey'),
                  makeData(artsobsFit, 'artsobs'),
                  makeData(ISDM, 'ISDM'),
                  makeData(ISDMbias, 'ISDMbias'))

ggplot(plotData, aes(x = mean, y = variable, col = model)) +
  geom_point(position=position_dodge(width=0.2)) +
  geom_errorbar(aes(x = mean, y = variable, xmin = min, xmax = max),  
                position=position_dodge(width=0.2), width = 0) +
  geom_vline(xintercept = 0, lty = 2) +
  theme_minimal() +
  theme(text = element_text(family = f1),
        strip.text = element_text(family = f1, size = 12, face = "italic"),
        plot.tag = element_text(angle = 90, hjust = 0.5),
        plot.tag.position = c(-0.03, 0.45))

```

We define a function that will do species-specific predictions for the ISDM with a bias field, and save the species predictions, since these take a little time to compute.

```{r, prediction function}
predict_species <- function(model, species, predict_data){
  sharedfield <- predict(model, 
                         data = predict_data,
                         spatial = TRUE, 
                         fun = 'linear',
                         species = species,
                         n.samples = 100)
  file_name <- paste0("results/sharedfield_", species, ".rds")
  saveRDS(sharedfield, file_name)
  return(sharedfield)
}
```

```{r, predict spatial fields for all four fish species, eval = run_prediction}
predData <- fm_pixels(mesh = mesh, mask = norway.poly, dims = c(450, 450))
prediction_list <- list()
for(fish in fishes) {
  prediction_list[[fish]] <- predict_species(
    model = ISDMbias,
    predict_data = predData,
    species = fish
  )
}
```

Once we have the predictions, we can make some plots. We similarly define a function that makes a plot for one species, and then run this for all four species.

```{r, plotting shared fields from the fish model}
NO <- st_boundary(st_as_sf(norway.poly))

plot_preferences <- list(scale_color_distiller(palette = "BrBG", direction = 1), 
                         coord_sf(),
                         xlab(""), ylab(""),
                         scale_x_continuous(breaks = c(5, 25)),
                         scale_y_continuous(breaks = c(60, 68)),
                         theme_minimal(),
                         theme(text = element_text(family = f1),
                               title = element_text(family = f1, size = 10, face = "italic"),
                               legend.key.height = unit(0.3, "cm"), 
                               legend.title = element_blank(),
                               legend.position = "bottom")
)

plot_species <- function(predictions, species_to_plot, plot_preferences){
  
  p <- ggplot() +
    gg(predictions$speciesPredictions[[species_to_plot]],
       aes(col = mean)) +
    gg(NO, lwd = 0.8) +
    labs(title = sub("_", " ", species_to_plot)) +
    plot_preferences 
  return(p)
}
```

```{r, plot spatial fields for all four fish species, eval = TRUE,fig.width=8, fig.height=5}
plot_list <- list()
for(fish in fishes) {
  plot_list[[fish]] <- plot_species(
    predictions = readRDS(paste0("results/sharedfield_", fish, ".rds")),
    species = fish,
    plot_preferences = plot_preferences)
}

patchwork::wrap_plots(plot_list, nrow = 1)
ggsave("figures/fishplot_four_species.pdf", height = 3*2, width = 8*2)
ggsave("figures/fishplot_four_species.png", height = 3*2, width = 8*2)
```

And finally we predict and plot the bias field, which is shared between all the fish, as it describes the human sampling more than the distribution of the fish.

```{r, plotting bias field from the four fish model, eval = TRUE,fig.width=8, fig.height=5}
fish_biasfield <- predict(ISDMbias, 
                          data = predData,
                          biasfield = TRUE,
                          fun = 'linear', 
                          n.samples = 100) 

saveRDS(fish_biasfield, "results/ISDMbiasfield.rds")
fish_biasfield <- readRDS("results/ISDMbiasfield.rds")

ggplot() +
  gg(fish_biasfield$biasFields$artsobs,
     aes(col = mean)) +
  gg(NO, lwd = 0.8) +
  labs(title = "Bias field") +
  plot_preferences

ggsave("figures/ISDMbiasfield.pdf", width = 3*2, height = 3*2)
```

# Session info and runtime

```{r, echo = FALSE}
end_time <- Sys.time()
knit_time <- round(difftime(end_time, start_time, units = "hours"), 2)
```

This document took `r knit_time` hours to compile.

```{r, echo = FALSE}
sessioninfo::session_info(pkgs = "loaded") |> 
  details::details(
    summary = 'Current session info',
    open    = TRUE
  )
```
