---
title: "Supporting information for 'The Point Process Framework for Integrated Modelling of Biodiversity Data'"
author: "Kwaku Peprah Adjei, Philip Mostert, Jorge Sicacha Parada, Emma Skarstein, Robert B. O'Hara"
output: 
  pdf_document
header-includes:
  - \geometry{top=1in}
  - \usepackage{titling}
  - \pretitle{\begin{flushleft}\Huge\bfseries}
  - \posttitle{\end{flushleft}}  
  - \preauthor{\begin{flushleft}\Large}
  - \postauthor{\end{flushleft}}  
  - \predate{\begin{flushleft}\large}
  - \postdate{\end{flushleft}}
---

Last updated: `r format(Sys.Date(), "%B %d %Y")` at `r format(Sys.time(), "%H:%M:%S")`.

```{r, echo = FALSE}
# Run model, or load from saved version? (and same for predictions)
run_model <- TRUE
run_prediction <- TRUE
```

```{r, echo = FALSE}
start_time <- Sys.time()
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.fullwidth = TRUE)
```

# Setup

These are the packages we will be using:

```{r, loading packages, message = FALSE}
#devtools::install_github("PhilipMostert/PointedSDMs")
library(PointedSDMs) # model fitting
library(ggplot2)     # plotting
library(raster)      # ???? Model fitting fails if I exclude this
library(mapproj)     # map options for plotting
library(INLA)        # functions for specifying mesh
library(dplyr)       # data handling
library(sf)          # spatial stuff
library(showtext)    # font for plot
library(patchwork)   # combining figures
```



# Downloading data

Before you run this file, make sure you have the following files in the given locations:

- `data/environmental_covariates.rds` 
  - should be in the data folder already.
- `data/Norwegian_lakes.rds`
  - can be downloaded from https://bird.unit.no/resources/9b27e8f0-55dd-442c-be73-26781dad94c8/content, click on “Innhold”-tab at the bottom of the page and select "Norwegian_lakes.rds".
- `data/artsobs_clean.rds`
  - created from running the script `R/data_preparation.R`, see further instructions there.
- `data/survey_clean.rds`
  - also created from running the script `R/data_preparation.R`, see further instructions there.



# Loading Norway map and making mesh

We begin by making a spatial mesh out of a map of Norway.

```{r, norway map stuff, warning = FALSE, message = FALSE, eval = FALSE, echo = FALSE}
#norway <- ggplot2::map_data("world", region = "Norway(?!:Svalbard)")
#norway <- setdiff(norway, dplyr::filter(norway, subregion == "Jan Mayen"))
#norway_sf <- st_as_sf(norway, coords = c("long", "lat"), crs = proj)
```


```{r, norway map, warning = FALSE, message = FALSE}
proj <- '+proj=utm +zone=32 +datum=WGS84 +units=m +no_defs'
norwayfill <- maps::map("world", "norway", fill=TRUE, plot=FALSE, 
                  ylim=c(58,72), xlim=c(4,32))
IDs <- sapply(strsplit(norwayfill$names, ":"), function(x) x[1])
norway.poly <- maptools::map2SpatialPolygons(norwayfill, IDs = IDs, 
                                             proj4string = CRS(proj))
```

Adjusting the mesh to be coarser is the easiest way to decrease the run-time for the model. With the following mesh, the model fitting takes a little more than 1 hour for us, but feel free to change the `max.edge` or `cutoff` to get a coarser mesh.

```{r, make small mesh, eval = TRUE}
mesh <- inla.mesh.2d(boundary = inla.sp2segment(norway.poly), 
                     cutoff = 0.3, # smallest allowed distance between points
                     max.edge = c(6, 3), # decrease this for more int. points
                     #max.edge = c(3, 1),
                     offset = c(1, 1),
                     crs = st_crs(proj))
plot(mesh)
```


# Setting up covariate data

Next we load the environmental data, which will be used as covariates.

```{r, loading covariates}
covariates_raw <- readRDS("data/environmental_covariates.RDS")

covariates <- covariates_raw %>% 
  # Log-transform area of lake
  dplyr::mutate(log_area = log(area_km2)) %>% 
  # Remove some uninformative variables
  dplyr::select(-c(ebint, no_vatn_lnr, eb_waterregionID))

# Choose from 
# "decimalLatitude", "decimalLongitude",
# "log_area", "perimeter_m", "distance_to_road", 
# "eurolst_bio10", "catchment_area_km2", "SCI", "HFP"

Use <- c("log_area", "eurolst_bio10", "SCI")

cov_pixel <- SpatialPixelsDataFrame(
  points = covariates[,c("decimalLongitude","decimalLatitude")],
  data = covariates[,Use], 
  proj4string = CRS(proj), 
  tol = 0.340571)

# Scale covariates and convert to terra::rast
cov_raster <- scale(terra::rast(cov_pixel))
```

# Observation data

For this model, we have two observation sets, one which is downloaded from GBIF and one that is a survey dataset (see separate document for download instructions).

```{r, loading observation data}
fishes <- c("Esox_lucius", "Perca_fluviatilis", "Salmo_trutta", "Salvelinus_alpinus")

survey <- readRDS("data/survey_clean.rds") %>% 
  filter(species %in% fishes)
artsobs <- readRDS("data/artsobs_clean.rds") %>% 
  filter(species %in% fishes)
```

We can plot the observed data points:

```{r, plot observations}
showtext_auto()
f1 <- "Open sans"
font_add_google(f1, f1)

norway <- ggplot2::map_data("world", region = "Norway(?!:Svalbard)")
norway <- setdiff(norway, dplyr::filter(norway, subregion == "Jan Mayen"))

p_artsobs <- ggplot(artsobs, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_polygon(data = norway, aes(long, lat, group = group), 
               color="grey80", fill = "grey95") + 
  geom_point(color = "darkorange2", size = 0.5, alpha = 0.3) +
  facet_wrap(~species, nrow = 1, 
             labeller = labeller(species = function(string) sub("_", " ", string))) +
  coord_map() +
  labs(tag = "Citizen science data") +
  theme_minimal() +
  theme(text = element_text(family = f1),
        strip.text = element_text(family = f1, size = 12, face = "italic"),
        plot.tag = element_text(angle = 90, hjust = 0.5),
        plot.tag.position = c(-0.03, 0.45),
        legend.position = "none",
        axis.title = element_blank(),
        axis.text.x = element_blank())

p_survey <- ggplot(survey %>% filter(occurrenceStatus == 1), 
       aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_polygon(data = norway, aes(long, lat, group = group), 
               color="grey80", fill = "grey95") + 
  geom_jitter(color = "darkcyan", size = 0.5, alpha = 0.3) +
  facet_wrap(~species, nrow = 1) +
  coord_map() +
  xlab("Longitude") +
  ylab("Latitude") +
  labs(tag = "Survey data") +
  theme_minimal() +
  theme(text = element_text(family = f1),
        plot.tag = element_text(angle = 90, hjust = 0.5),
        plot.tag.position = c(-0.03, 0.45),
        legend.position = "none",
        axis.title = element_blank(),
        strip.text = element_blank(),
        plot.margin = margin(l = 30))

p_artsobs / p_survey

ggsave("figures/presence_points.pdf", height = 5, width = 8)
```

# Joint model for four fish species

For the presence/absence survey data, we use a Bernoulli distribution, where the presence probability for species $j\in \{\textit{Salmo trutta, Perca fluviatilis, Esox lucius, Salvelinus alpinus}\}$ depends on some covariates $x(s)$, along with a spatial field $\xi_{j}(s)$:

$$
\begin{aligned}
    Y_{PA, j}(s_i) &\sim \text{Bernoulli}(p_{PA, j}(s_i)) \\
    \text{cloglog}(p_{PA, j}(s_i)) &= \alpha_{PA, j} + x(s_i)^T\beta_j + \xi_{j}(s_i).
\end{aligned}
$$

The presence-only data is fitted with a Poisson point process model, where the intensity depends on the same covariates $x(s)$ and the same spatial field $\xi_{j}(s)$, plus an additional spatial field $\xi_{\text{bias}}(s)$ that is unique to the citizen science data, but shared across all fish species:

$$
\begin{aligned}
    Y_{PO, j}(s_i) &\sim \text{Poisson}(e^{\eta_{PO, j}(s_i)}) \\
    \eta_{PO, j}(s_i) &= \alpha_{PO, j} + x(s)^T\beta_j + \xi_{j}(s_i) + \xi_{\text{bias}}(s_i).
\end{aligned}
$$

So in summary, for this model we have:

  - one bias field, based on CS data and common across all species
  - four shared fields, shared across the data sets (survey/citizen science), but separate for each fish species.
  
Since we have two data sets and four species, that means that we in total have eight sub-models.

We first prepare the model using the `intModel` function.
  
```{r, setting up fish model, warning=FALSE}
fish_model_setup <- intModel(
  survey,                              # Survey data
  artsobs,                             # Citizen science data
  spatialCovariates = cov_raster,      # Covariates
  speciesName = "species",             # The column containing species name
  speciesSpatial = "copy",             # Copy species fields across data
  Coordinates = c("decimalLongitude", "decimalLatitude"), # Name of coords
  responsePA = "occurrenceStatus",     # Name of response column
  pointsSpatial = NULL,                # NULL since we use speciesSpatial
  Mesh = mesh,                         # inla mesh object
  Projection = proj)                   # CRS for points and covariates
```

This model has the species specific spatial fields, but we also want a bias field that is shared across the species. We add this using `$addBias`.

```{r, add bias field, warning=FALSE}
fish_model_setup$addBias("artsobs")
```

For the species specific fields, the default in PointedSDMs is that these are allowed to be different up to a scaling factor (named beta in INLA) across the data sets. That means that for instance the trout-specific spatial field for the citizen science data set is equal to beta times the trout-specific spatial field for the survey data. In practice, this has to do with the copy-option in INLA. But in our model we want them to be the same, not to vary by a factor. So we manually change this using the `$changeComponents` function, by setting `hyper = list(beta = list(fixed = TRUE))` for each of the four citizen science fields.

```{r, customizing fish model, warning=FALSE}
fish_model_setup$changeComponents(
  'Perca_fluviatilis_artsobs_spatial(
      main = geometry, 
      copy = "Perca_fluviatilis_survey_spatial", 
      hyper = list(beta = list(fixed = TRUE)))')

fish_model_setup$changeComponents(
  'Salmo_trutta_artsobs_spatial(
      main = geometry, 
      copy = "Salmo_trutta_survey_spatial", 
      hyper = list(beta = list(fixed = TRUE)))')

fish_model_setup$changeComponents(
  'Salvelinus_alpinus_artsobs_spatial(
      main = geometry,
      copy = "Salvelinus_alpinus_survey_spatial",
      hyper = list(beta = list(fixed = TRUE)))')

fish_model_setup$changeComponents(
  'Esox_lucius_artsobs_spatial(
      main = geometry, 
      copy = "Esox_lucius_survey_spatial", 
      hyper = list(beta = list(fixed = TRUE)))')
```


We may look at which terms are included in each of the eight sub-models by calling `$updateFormula` with the data sets as the arguments.

```{r, view formulas, warning=FALSE}
fish_model_setup$updateFormula(datasetName = "survey")
fish_model_setup$updateFormula(datasetName = "artsobs")
```

Finally, we actually fit the model using `fitISDM`. 

```{r, fitting fish model, warning=FALSE, eval = run_model}
fish_model <- fitISDM(fish_model_setup, 
                      options = list(
                        control.inla = list(int.strategy = 'eb', cmin = 0),
                        safe = TRUE,
                        inla.mode = 'experimental'))
```

We may then examine the model summary and save the model for future use.

```{r, view and save model, warning=FALSE, eval = run_model}
summary(fish_model)

saveRDS(fish_model, "results/fish_model.rds")
```


```{r, load model, warning=FALSE, echo = FALSE}
fish_model <- readRDS("results/fish_model.rds")
```

# Predictions and plots

Once the model has been fit, we can look at the predictions from the species-specific shared fields and the bias field.

We define a function that will do species-specific predictions, and save the species predictions, since these take a little time to compute.

```{r, prediction function}
predict_species <- function(model, species, mask, mesh){
  sharedfield <- predict(model, 
                         mesh = mesh, 
                         mask = mask, 
                         format = 'sp',
                         spatial = TRUE, 
                         fun = 'linear', 
                         species = species,
                         n.samples = 1000)
  file_name <- paste0("results/sharedfield_", species, ".rds")
  saveRDS(sharedfield, file_name)
  return(sharedfield)
}
```

```{r, predict spatial fields for all four fish species, eval = run_prediction}
prediction_list <- list()
for(fish in fishes) {
  prediction_list[[fish]] <- predict_species(
    model = fish_model,
    mesh = mesh,
    mask = norway.poly,
    species = fish
  )
}
```

Once we have the predictions, we can make some plots. We similarly define a function that makes a plot for one species, and then run this for all four species.

```{r, plotting shared fields from the fish model}
plot_preferences <- list(scale_fill_distiller(palette = "BrBG", direction = 1), 
                         coord_map(),
                         xlab(""), ylab(""),
                         theme_minimal(),
                         theme(text = element_text(family = f1),
                               title = element_text(family = f1, size = 10, face = "italic"),
                               legend.key.height = unit(0.3, "cm"), 
                               legend.title = element_blank(),
                               legend.position = "bottom")
)

plot_species <- function(predictions, species_to_plot, plot_preferences){
  p <- ggplot() +
    geom_polygon(data = norway, aes(long, lat, group = group), 
               color="grey80", fill = "grey95") + 
    gg(predictions$speciesPredictions[[species_to_plot]]) +
    labs(title = sub("_", " ", species_to_plot)) +
    plot_preferences 
  return(p)
}
```

```{r, plot spatial fields for all four fish species, eval = TRUE}
plot_list <- list()
for(fish in fishes) {
  plot_list[[fish]] <- plot_species(
    predictions = readRDS(paste0("results/sharedfield_", fish, ".rds")),
    species = fish,
    plot_preferences = plot_preferences)
}

patchwork::wrap_plots(plot_list, nrow = 1)
ggsave("figures/fishplot_four_species.pdf", height = 3, width = 8)
```

And finally we predict and plot the bias field, which is shared between all the fish, as it describes the human sampling more than the distribution of the fish.

```{r, plotting bias field from the four fish model, eval = TRUE}
fish_biasfield <- predict(fish_model, 
                          mesh = mesh, 
                          mask = norway.poly, 
                          format = 'sp',
                          biasfield = TRUE,
                          fun = 'linear', 
                          n.samples = 1000) 

saveRDS(fish_biasfield, "results/fish_biasfield.rds")
fish_biasfield <- readRDS("results/fish_biasfield.rds")

ggplot() +
  geom_polygon(data = norway, aes(long, lat, group = group), 
               color="grey80", fill = "grey95") + 
  gg(fish_biasfield$biasFields$artsobs) +
  labs(title = "Bias field") +
  plot_preferences

ggsave("figures/fishplot_biasfield.pdf", width = 3, height = 3)
```



```{r, echo = FALSE}
end_time <- Sys.time()
knit_time <- round(difftime(end_time, start_time, units = "hours"), 2)
```

This document took `r knit_time` hours to compile.
