---
title: "Integrated lakefish"
output: html_document
date: `r Sys.Date()`
---

As a start, I think we can try to redo (this analysis)[https://htmlpreview.github.io/?https://github.com/emmaSkarstein/Effort_field_project/blob/main/R/Mini_test.html], but using Philip's `PointedSDMs`. So far, I have copied the data loading lines. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(PointedSDMs)
library(ggplot2)
library(raster)
library(INLA)
library(dplyr)

library(sf) # for sf objects
library(plyr)
library(ggmap) # also for nice maps
library(maps)
library(RColorBrewer)
library(INLA)
library(PointedSDMs)
library(sp)
library(spatstat)
library(maptools)
library(reshape2)
library(rgeos)
library(fields)
library(viridis)
library(colorspace)
library(patchwork)
```

```{r}
source("functions.R")
```

## Downloading necessary data 
I have uploaded all the cleaned data to the "data"-folder, except for the lake polygons which are too large to include on Github. But they are easily available here: 

- **Lake polygons for Norway:** Go to https://bird.unit.no/resources/9b27e8f0-55dd-442c-be73-26781dad94c8/content (click on "Innhold"-tab at the bottom of the page to download only selected sets of lakes). The object name should be Norwegian_lakes.rds, and it should be placed in a "data" folder on the top level (the same level as the R-project).

## Loading Norway map



I think `PointedSDMs` has a function for this, so probably rewrite some of the below.

```{r, Norway map}
# MAP ---------------------------------------------------------------------------------------------------------
norway <- ggplot2::map_data("world", region = "Norway(?!:Svalbard)")
norway <- setdiff(norway, dplyr::filter(norway, subregion == "Jan Mayen"))
Projection <- CRS("+proj=longlat +ellps=WGS84")
norwayfill <- maps::map("world", "norway", fill=TRUE, plot=FALSE, 
                  ylim=c(58,72), xlim=c(4,32))
IDs <- sapply(strsplit(norwayfill$names, ":"), function(x) x[1])
norway.poly <- maptools::map2SpatialPolygons(norwayfill, IDs = IDs, 
                                   proj4string = Projection)
```

```{r, make mesh}
# INTEGRATION STACK --------------------------------------------------------------------------------------------
Meshpars <- list(cutoff=0.08, max.edge=c(0.6, 3), offset=c(1,1))
Mesh <- MakeSpatialRegion(data=NULL, bdry=norway.poly, 
                                       meshpars=Meshpars,
                                       proj = Projection)


mesh <- inla.mesh.2d(boundary = inla.sp2segment(PA), 
                     cutoff = 0.2,
                     max.edge = c(0.1, 0.24), 
                     offset = c(0.1, 0.4))

mesh$crs <- proj

plot(Mesh$mesh)
```
## Setting up observation and environmental data

Next we load the observations as well as the environmental data.

```{r, loading data and covariates}
# LOADING DATA AND COVARIATES ---------------------------------------------------------------------------------

# Covariates
covariateData <- readRDS("data/environmental_covariates.RDS")
covariateData <- covariateData[complete.cases(covariateData$decimalLatitude,
                                              covariateData$decimalLongitude,
                                              covariateData$area_km2,
                                              covariateData$HFP),]
covariateData <- covariateData %>% mutate(log_area = log(area_km2)) %>% 
  select(-c(ebint, no_vatn_lnr, eb_waterregionID))

head(covariateData)

# Choose from 
# "decimalLatitude", "decimalLongitude",
# "area_km2", "perimeter_m", "distance_to_road", 
# "eurolst_bio10", "catchment_area_km2", "SCI", "HFP"
Use <- c("decimalLongitude","decimalLatitude", "log_area", #"perimeter_m", 
         "eurolst_bio10", "SCI")

Covariates <- SpatialPointsDataFrame(coords = covariateData[,c("decimalLongitude","decimalLatitude")],
                                     data = covariateData[,Use], 
                                     proj4string = Projection)
Covariates@data <- data.frame(apply(Covariates@data, 2, scale))  # scale the covariates

# Observations
Data_survey_df <- readRDS("data/survey_clean.rds")
Data_survey <- SpatialPointsDataFrame(coords = Data_survey_df[,c("decimalLongitude","decimalLatitude")], 
                                      data = Data_survey_df[,c("occurrenceStatus","species")],
                                      proj4string = Projection)

Data_artsobs_df <- readRDS("data/artsobs_clean.rds")
Data_artsobs <- SpatialPointsDataFrame(coords = Data_artsobs_df[,c("decimalLongitude","decimalLatitude")], 
                                       data = Data_artsobs_df[,c("occurrenceStatus","species")],
                                       proj4string = Projection)



# Separating by species:
perch_survey_df <- filter(Data_survey_df, grepl('Perca fluviatilis', species))
trout_survey_df <- filter(Data_survey_df, grepl('Salmo trutta', species))
char_survey_df <- filter(Data_survey_df, grepl('Salvelinus alpinus', species))
pike_survey_df <- filter(Data_survey_df, grepl('Esox lucius', species))

perch_artsobs_df <- filter(Data_artsobs_df, grepl('Perca fluviatilis', species))
trout_artsobs_df <- filter(Data_artsobs_df, grepl('Salmo trutta', species))
char_artsobs_df <- filter(Data_artsobs_df, grepl('Salvelinus alpinus', species))
pike_artsobs_df <- filter(Data_artsobs_df, grepl('Esox lucius', species))


MakeSpDF <- function(df){
  Projection <- CRS("+proj=longlat +ellps=WGS84")
  sp_df <- SpatialPointsDataFrame(coords = df[,c("decimalLongitude","decimalLatitude")], 
                                  data = df[,c("occurrenceStatus","species")],
                                  proj4string = Projection)
  sp_df
}

perch_survey <- MakeSpDF(perch_survey_df)
trout_survey <- MakeSpDF(trout_survey_df)
char_survey <- MakeSpDF(char_survey_df)
pike_survey <- MakeSpDF(pike_survey_df)

perch_artsobs <- MakeSpDF(perch_artsobs_df)
trout_artsobs <- MakeSpDF(trout_artsobs_df)
char_artsobs <- MakeSpDF(char_artsobs_df)
pike_artsobs <- MakeSpDF(pike_artsobs_df)
```
